On the spatial aggregation of ecosystem condition indicators

### 2025-01-15 manuscript/manuscript.qmd
Writing

In ecosystem accounting (EA), fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
This simplification is necessary to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers.
The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs.
In ecosystem condition accounting (ECA) specifically, the aggregation of both variables and their normative indicators is common practice and raises the issue of aggregation bias when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
The first two sources of error are usually explicitly addressed and managed, but the last case of aggregation bias is often overlooked.
In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (a thematic aggregation), but in this paper we are discussing spatial aggregation only.

Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
They have a spatial component, tying them to concrete areas, which can be of varying size, from single pixels to entire regions.
Variables can be spatially aggregated, for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
Variables can also be normalised and turned into normative indicators for ecosystem condition.
The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
The order of these steps can also change.
Variable can be scaled using at least two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
Variables can be truncated to produce a bound indicator scale between 0 and 1.
Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
When a variable is normalised in this way, its interpretation changes from descriptive to normative.
The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference levels.
The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).

Similar to variables, the reference values used for normalising the variables also have a spatial assignment, but it can be different from that of the variable.
For example, variable values may exist for unique 10×10m grid cells, but the reference values may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire EAA (e.g. a natural zero).
A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (@fig-pathways).

@@

Normalisation of ecosystem condition indicators, as defined in @fig-term, serves multiple purposes.
Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
<!--# This normative characteristic is an essential part of assessing ecosystem condition. --> Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa.
This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
Transformation can sometimes also have this effect, e.g. sigmoid transformations.
<!--# This is often what we want (but not always), but not always something we are aware that we are doing. -->One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition.
(An exception if the WFD; see @sec-pti).
Therefore, this way of aggregation summarises the estimated ecosystem condition (which is generally what we want), and not the variable itself.
In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
This is commonly referred to as the reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.

@@

>>>>

Ecosystem accounting (EA), and other types of environmental assessments, are generally aimed at communicating major changes and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something actionable.
However, spatial aggregation can and will often change the interpretation of a metric and the descriptive properties and normative claims one can make based on it.
Without a good enough understand of the pitfalls in spatial aggregation of information, is can also lead to unintended and undetected systematic bias, as well as downright logical errors.
Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (ref), as well as to the spatial aggregation of relatively unmodified or raw variables, there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting.
This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
We pay specific attention to the phenomenon of aggregation bias introduce by @rastetter1992.

[^1]: Words in bold are explained in Table 1

### SEEA EA

SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services.
We will focus on the SEEA EA condition accounts in this paper.
In these accounts, information about ecosystem condition is made spatially explicit by assigning **variable** or **indicator** values to individual ecosystem assets.
Unless the underlying data are wall-to-wall maps, for example from satellite images, then this process of assigning values to ecosystem assets already here requires some spatial manipulations in the form of aggregation, dissaggregation, extrapolation or interpolation (ref Ewert 2011?), of which averaging or summing are the most common methods.
Subsequently, information from ecosystem assets may be spatially aggregated to larger areas, such as the entire ecosystem accounting area.
When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.

### Indicators vs. variables

SEEA EA distinguishes between indicators and variables.
Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
They are typically either in units or percentages or fractions, or as raw biophysical units, such as biomass, density, or cover.
Indicators on the other hand, are always on a scale between 0 and 1, where 0 represent the worst condition, and 1 represents the best condition.
Sometimes indicators are calculated directly, but usually indicators are made by **normalising** variables based on two or more **reference levels**.
Although also variables may have normative properties (due to choices about how they are chosen (ref), defined, aggregated or presented) (Alain et al), the normalisation of variables into indicators makes indicators explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition.
The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
The order of these steps can also change.

Variables can be **scaled** using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
Variables can be **truncated** to produce a bound indicator scale between 0 and 1.
Sometimes variables are **transformed** to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.

| Term             | Definition |
|------------------|------------|
| Metric           |            |
| Variable         |            |
| Indicator        |            |
| Reference levels |            |
| Scaling          |            |
| Transformation   |            |
| Truncation       |            |
| Normalisation    |            |
|                  |            |
|                  |            |

: Definition of terms

# Why to we spatially aggregate data in ecosystem condition accounting

Spatiall aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation[@allain2018].
Generally, spatial aggregation increases the tangibility of a metric, because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
This is necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g. nations) and when variables and indicators need to be presented in standardised tables.
Sometimes total aggregation of a variable or indicator is required for the metric to actually fit its purpose and become relevant to what it is designed to describe.
For example, a hypothetical metric "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
The reason and need for aggregation may therefore originate in the original idea for the metric.
Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation [@allain2018].

A minimum level of aggregation may be decided on based on the uncertainties in the data.
High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use [@lyytimäki2013], but these values seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons [@allain2018].
For example, ecosystem condition metrics can be visualised on a map, but when both the the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
Similarly, one may chose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the aggregated values between these areas.

However, spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
This smoothing out makes the data and its inherent information less suitable for other uses, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts.

### Matching the scale of variables and reference levels

Similar to variables, the reference levels used for normalising variables also have a spatial assignment, but it can be different from that of the variable.
For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g. a natural zero).
Variables should not be normalised based on reference levels unless they are at a comparable scale.
Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the **reference condition** (the optimal condition), then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with deteriorated condition simply due to natural variation in where the animals are at any give time.
In addition, averaging the indicator values across sub-regions would not produce a regional indicator value that represents its true situation as being perfectly aligned with the reference condition, since truncation ensures that variable values above X~100~ (assuming X~100~ \> X~0~) means we are only averaging the deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.

In conclusion, there are several requirements in the SEEA EA that make spatial aggregation necessary, but spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
Readers are referred to @allain2018 for a review of different reasons for spatially aggregating indicators.

@@

Normalisation of ecosystem condition indicators, as defined in @fig-term, serves multiple purposes.
Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
SEEA EA encourages normalisation, but other environmental assessment framework leaves more of these kinds of value judgments to the end users (ref?).

Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
Transformation can sometimes also have this effect, e.g. sigmoid transformations.
One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition (an exception if the WFD; see @sec-pti).
Therefore, this way of aggregation summarises the estimated ecosystem condition over an area (which is generally what we want), and not the variable itself.
In other words, normalising variables facilitate the spatial aggregation of our ecosystem condition estimates (i.e. the indicator values).

Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation (ref).
This is commonly referred to as *the* reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.

@@

# Aggregation bias

The spatial aggregation of data can lead to biases arising from a number of sources.
For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
However, we are here mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992. They showed that when we take the average of two variables, each of which have associated attribute values (e.g. indicator values), then the attribute of the average variable value differs from the average of the attributes of the original variables.
This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.

Also, in ecosystem condition accounting (ECA) specifically, the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1; @fig-agg2) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.

A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (@fig-pathways).

# Indicator developer subjectivity

Indicator developer subjectivity is a big factor as the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
This paper is aimed at providing some recommendations for how to choose the right spatial aggregation method for variables and indicators, and how to report these choices.
This will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.

The choice of aggregation function (e.g. weighted arithmetic mean, worst-rule, or others) also influences the degree of compensation, i.e. the degree to which good condition in one place can compensate for bad condition somewhere else (ref WFD?).
Therefore, spatial aggregation can introduce substantial subjectivity from the side of the indicator developers, which also limits the different ways that a metric can be used and interpreted by indicator users (the target audience).
In other words, spatial aggregation tasks indicator developers with defining narratives for underlying variables, and it is therefor a normative and value-laden operation.

Something about sometimes spatial aggregation is an essential part of the indicator design ---

In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e. thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e. horizontal) aggregation only.
